---
title: "Plot Extras"
author: "Cole Baril"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Overview

This vignette documents the "plot extras" functionality in the package, including:

* Adding social media captions to plots and tables
* Readable numeric formatting with commas
* Colour palettes and scales for ggplot2

These functions enhance visualization without altering the core plotting logic.

---

## 1. Adding a caption with `add_caption_cwb()`

`add_caption_cwb()` allows you to add a social media caption to ggplot2 plots. It supports GitHub and Bluesky usernames and optionally includes a data source.

### Example: ggplot2 plot

```{r}
library(ggplot2)
library(trashpanda)

p <- ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  labs(title = "Displacement vs Highway MPG")

p + add_caption_cwb(type = "plot", github_username = "colebaril", include_data_source = TRUE, data_source = "mpg dataset")
```

---

## 2. Numeric formatting with `add_commas()`

`add_commas()` formats large numbers with commas for easier readability.

```{r}
x <- c(1000, 25000, 1234567)
add_commas(x)
#> "1,000" "25,000" "1,234,567"
```

This is particularly useful for labeling axes or tables where numeric readability is important.

---

## 3. Colour palettes and scales

The package includes curated palettes for different visual moods. Examples include `arcane_flame`, `enchanted_forest`, `mystic_ocean`, and more.

### Using a discrete palette

```{r}
ggplot(mpg, aes(class, fill = class)) +
  geom_bar() +
  scale_cwb(palette = "discrete_21", type = "d", aesthetics = "fill")
```

### Using a continuous palette

```{r}
ggplot(mpg, aes(displ, hwy, colour = hwy)) +
  geom_point(size = 3) +
  scale_cwb(palette = "mystic_ocean", type = "c", aesthetics = "colour")
```

### Available palettes

* arcane_flame
* enchanted_forest
* mystic_ocean
* eldritch_night
* golden_parchment
* discrete_21, discrete_31, discrete_35
* spotify_pal
* halloween
* haunted_forest
* ghostly_glow
* blood_moon

---

## Summary

This vignette highlights utility functions that enhance plots and tables:

* `add_caption_cwb()` for social media and data source captions
* `add_commas()` for readable numeric labels
* `scale_cwb()` for applying curated palettes

These extras integrate seamlessly with `ggplot2` and `gt` objects to streamline polished figure creation.

