---
title: "Plot Themes"
author: "Cole Baril"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Often when creating plots and visuals I like them to be highly customized. I found I routinely came back to a few themes. Thus, I created several functions to make applying themes to plots and adding elements to plots a little easier. 

## `theme_cole()`

`theme_cole()` is a custom `ggplot2` theme designed to provide a restrained, manuscript-like aesthetic. The goal was not heavy ornamentation, but a clean default that:

* Emphasizes titles and axes
* Reduces unnecessary visual noise
* Provides fine-grained control over axis lines
* Supports dark mode and transparent backgrounds

---

### Design principles

The theme is built on top of `theme_minimal()` and modifies only a focused set of elements:

* Typography: bold titles, slightly larger legend text
* Layout: increased margins and centered titles
* Axes: explicit control over which axes are shown
* Backgrounds: light, dark, or transparent

The intent is that the theme works across many plot types without requiring per-plot tweaking.

---

### Basic usage

```{r}
library(ggplot2)
library(palmerpenguins)
library(trashpanda)

p <- ggplot(penguins, aes(flipper_length_mm, bill_length_mm)) +
  geom_point()

p + theme_cole()
```

By default:

* Grid lines are retained
* Axis lines appear only on the bottom and left
* Light background is used

---

### Axis control

A central feature of `theme_cole()` is explicit axis control via `show_axis_lines`.

Valid values are:

* `"bottom"`, `"top"`, `"left"`, `"right"`
* `"all"`
* `"none"`

#### Example: bottom and left only

```{r}
p + theme_cole(show_axis_lines = c("bottom", "left"))
```

#### Example: all axes

```{r}
p + theme_cole(show_axis_lines = "all")
```

When `top` or `right` axes are requested, secondary axes are automatically created using `dup_axis()`.

#### Example: no axes

```{r}
p + theme_cole(show_axis_lines = "none")
```

This removes axis lines, ticks, text, and panel borders entirely.

---

### Grid control

Grid lines can be fully removed using `remove_grid = TRUE`.

```{r}
p + theme_cole(remove_grid = TRUE)
```

This is useful for presentation-style figures or plots intended to resemble hand-drawn charts.

---

### Dark mode

Setting `dark = TRUE` switches the theme to a dark background while preserving contrast and readability.

```{r}
p + theme_cole(dark = TRUE)
```

Dark mode changes:

* Plot and panel backgrounds
* Grid line colors
* Text, axis, and legend colors

Axis line color is automatically adjusted to white.

---

### Transparent backgrounds

For slides, overlays, or compositing figures, a transparent background can be requested:

```{r}
p + theme_cole(transparent = TRUE)
```

This affects:

* Plot background
* Panel background
* Legend and strip backgrounds

Note that transparency applies only in light mode.

---

### Typography and sizing

The theme exposes standard sizing parameters:

* `base_size`: base font size (default 12)
* `base_family`: font family (default `"sans"`)

Titles are scaled relative to `base_size`:

* Plot title: `1.5 × base_size`
* Legend title: `1.2 × base_size`

This keeps proportions consistent across outputs.

---

### Example: publication-style scatter plot

```{r}
ggplot(penguins, aes(flipper_length_mm, bill_length_mm, fill = species)) +
  geom_point(shape = 21, size = 3) +
  labs(
    title = "Flipper Length vs Bill Length",
    subtitle = "Palmer Penguins dataset"
  ) +
  theme_cole(remove_grid = TRUE)
```

This configuration emphasizes data points and labels while minimizing distractions.

---

### Example: dark presentation slide

```{r}
ggplot(penguins, aes(flipper_length_mm, bill_length_mm, color = species)) +
  geom_point(size = 2) +
  labs(title = "Penguin Morphometrics") +
  theme_cole(dark = TRUE, remove_grid = TRUE)
```

---

### When to use `theme_cole()`

This theme works best when:

* You want consistent, restrained styling across figures
* Axis placement matters (e.g., publication or teaching figures)
* You need light, dark, and transparent variants without redefining themes

It is intentionally opinionated but avoids heavy stylistic elements so it can be layered with captions, annotations, and custom geoms.

---

### Summary

`theme_cole()` provides:

* A clean extension of `theme_minimal()`
* Explicit axis-line control
* Dark and transparent modes
* Sensible typography defaults

The implementation prioritizes clarity, reuse, and predictable behavior across plotting contexts.


