---
title: "Utilities"
author: "Cole Baril"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

## Overview

This vignette covers general utility functions included in the package:

* `cite_packages()` – generates formatted citations for loaded packages
* `load_quarto_theme()` – loads a precompiled Quarto CSS theme into your project

These utilities help manage reproducibility and presentation styling.

---

## 1. Citing Packages with `cite_packages()`

`cite_packages()` prints package citations in either R Markdown/Quarto style or plain text for the console. By default, it cites all currently loaded packages.

### Example: Markdown/Quarto citations

```{r, results = 'asis', comment = NA}
library(pacman)
# load some packages
p_load(trashpanda, ggplot2, dplyr, purrr)

cite_packages(format = "rmd")
```

Output is formatted for direct copy-paste into R Markdown or Quarto documents.

### Example: Plain text for console

```{r}
cite_packages(format = "text")
```

This prints citations in plain text, suitable for scripts or reports.

> Note: Packages without a registered citation will generate a warning.

---

## 2. Loading a Quarto theme with `load_quarto_theme()`

`load_quarto_theme()` copies a precompiled CSS file from the package into your project so you can use it in Quarto presentations with `revealjs`.

### Example: load the dark theme

```{r}
css_path <- load_quarto_theme(name = "dark")
# Use in Quarto YAML:
# css: styles/trashpanda-dark.css
```

### Example: load the light theme into a custom folder

```{r}
css_path <- load_quarto_theme(name = "light", dest_folder = "styles/custom", overwrite = TRUE)
```

This ensures consistent styling for slides without manually copying files.

---

## Summary

The utilities vignette demonstrates:

* How to create reproducible citations with `cite_packages()`
* How to integrate precompiled Trashpanda Quarto themes into your projects with `load_quarto_theme()`

These tools complement the visualization and workflow functions in the package, supporting reproducible research and presentation-ready outputs.
